Strong lensing selection effects

### 2022-08-07 paper/seleff.tex
started section 5

\citet{MVK09} carried out a thorough study of point 1): they quantified how the properties of a lens galaxy determine its probability of creating a lensing event with a point source.
In this work we revisit the \citet{MVK09} study, we expand it to the extended source case and, most importantly, we address point 2) as well: we simulate large populations of strong lenses using empirical models, simulate lensing surveys with realistic assumptions for the lens probability $\pdet$ and quantify the lensing bias under various scenarios.
We explore how the results change as a function of the scatter in mass parameters at fixed light, and as a function of source type: we simulate both galaxy and quasar sources.

@@

For each of the three intrinsic scatter scenarios, we drew a population of foreground galaxies covering XXX square degrees.
The expectation value of the number of galaxies in the corresponding volume, given the redshift and stellar mass cuts described in section \ref{ssub:mstarz}, is YYY.
%Then we generated a population of sources behind the lenses.
%We approximated the foreground galaxies as isolated: for each source, we only modelled the contribution to the lensing field generated by a single lens.
We approximated the foreground galaxies as isolated: when determining whether a galaxy acts as a strong lens, we only modelled the contribution to the lensing signal from the galaxy itself, and neglected that of the environment.
We discuss the possible implications of this approximation in \Sref{sect:discuss}.
For each lens, we determined its caustics relative to the highest source redshift, $\zsource=2.5$. We then placed sources randomly behind the lens. If at least one source fell within a circular region enclosing the caustics, we proceeded to compute its lensed images.

For the extended sources, we produced images with a pixel size of $0.1''$.
We applied a Moffat PSF with an FWHM of $0.2''$.
Finally, we assumed a background noise level such that an extended source with half-light radius $0.5''$ and an apparent magnitude in the absence of lensing of $\msource=25$ is detected with $S/N=10$.
%This is similar to the background noise expected for the Euclid Wide survey \citep{Sca++22}.
These values are close to those expected for the Euclid Wide survey \citep{Sca++22}.

We applied two different lens selection criteria.
First, we used the criterion described in section \ref{ssec:lensdetext}.
This resulted in a sample of XXX lenses.
Then, we added a cut on the Einstein radius of the lens: we only included in the lens sample those with $\tein > 0.5''$.
This additional cut is meant to approximate the difficulty in detecting strongly lensed images that lie very close to the lens centre: the number of known lenses with Einstein radius smaller than that cutoff value is very small, even in lens surveys that are based on high-resolution data \citep[see e.g.][]{Gar++22}.
With this more restrictive criterion, the number of strong lenses in the simulated sample is YYY.

\Fref{fig:lenspars} shows the distribution in the parameters of the foreground galaxies of the lens systems, compared to those of the general population.
\Fref{fig:sourcepars} shows the distribution in source properties.

@@

There are results.

>>>>

\citet{MVK09} carried out a thorough study of point 1): they quantified how the properties of a lens galaxy determine its probability of creating a lensing event with a point source.
In this work we revisit the \citet{MVK09} study, we expand it to the extended source case and, most importantly, we address point 2) as well: we simulate large populations of strong lenses using empirical models, simulate lensing surveys with realistic assumptions for the lens probability $\pdet$ and quantify the lensing bias under various scenarios.
%We explore how the results change as a function of the scatter in mass parameters at fixed light, of the efficiency of finding lenses with small image separation, and of source type: we simulate both galaxy and quasar sources.
%We explore how the results change as a function of the scatter in mass parameters at fixed light, and as a function of the efficiency of a lens survey at discovering small image separation lenses.
We explore how the results change as a function of the efficiency of a lens survey at discovering small image separation lenses, and of the scatter in mass parameters at fixed light.

@@

For each of the three intrinsic scatter scenarios, we drew a population of foreground galaxies covering $1000$ square degrees.
The expectation value of the number of galaxies in the corresponding volume, given the redshift and stellar mass cuts described in section \ref{ssub:mstarz}, is YYY.
%Then we generated a population of sources behind the lenses.
%We approximated the foreground galaxies as isolated: for each source, we only modelled the contribution to the lensing field generated by a single lens.
We approximated the foreground galaxies as isolated: when determining whether a galaxy acts as a strong lens, we only modelled the contribution to the lensing signal from the galaxy itself, and neglected that of the environment.
We discuss the possible implications of this approximation in \Sref{sect:discuss}.
For each lens, we determined its caustics relative to the highest source redshift, $\zsource=2.5$. We then placed sources randomly behind the lens. If at least one source fell within a circular region enclosing the caustics, we proceeded to compute its lensed images.

%\subsubsection{Lensed galaxies}

For the simulation with extended sources, we produced images with a pixel size of $0.1''$.
We applied a Moffat PSF with an FWHM of $0.2''$.
Finally, we assumed a background noise level such that an extended source with half-light radius $0.5''$ and an apparent magnitude in the absence of lensing of $\msource=25$ is detected with $S/N=10$.
%This is similar to the background noise expected for the Euclid Wide survey \citep{Sca++22}.
These values are close to those expected for the Euclid Wide survey \citep{Sca++22}.
%
We then applied the peak detection-based lens selection criterion introduced in section \ref{ssec:lensdefext} to find the strong lenses.

%First, we used the criterion described in section \ref{ssec:lensdefext}.
%This resulted in a sample of XXX lenses.
%Then, we added a cut on the Einstein radius of the lens: we only included in the lens sample those with $\tein > 0.5''$.
%This additional cut is meant to approximate the difficulty in detecting strongly lensed images that lie very close to the lens centre: the number of known lenses with Einstein radius smaller than that cutoff value is very small, even in lens surveys that are based on high-resolution data \citep[see e.g.][]{Gar++22}.
%With this more restrictive criterion, the number of strong lenses in the simulated sample is YYY.

%\subsubsection{Lensed quasars}

For the lensed quasars we did not simulate pixel-level data, but simply computed the observed magnitudes of the multiple images.
Following the definition of section \ref{ssec:lensdefpoint}, we included in the sample of strong lenses only systems with at least two images brighter than a limiting magnitude $m_{\mathrm{lim}}$.
We set $m_{\mathrm{lim}}=23.3$, which corresponds to the $10\sigma$ detection limit of the LSST in a single visit \citep{O+M10}. This is motivated by the fact that the quasar lenses are meant to simulate a sample assembled for the purpose of carrying out time delay measurements, which in turn require combining single-visit detections over many epochs.

%Both for the lensed galaxies and lensed quasars, we apply an additional selection based on the angular size of the Einstein radius.
Both for the lensed galaxies and lensed quasars, we apply an additional cut, by selecting only lenses with an angular Einstein radius larger than a limiting value $\theta_{\mathrm{Ein,min}}$.
This additional cut is meant to approximate the difficulty in detecting strongly lensed images with small separation, because of the contamination from the light of the lens galaxy.
We explored scenarios with different values of $\theta_{\mathrm{Ein,min}}$.

@@

\subsection{Galaxy-galaxy lenses}

\Fref{fig:lenspars} shows the distribution in the parameters of the foreground galaxies of the lens systems, compared to those of the general population, for the fiducial scatter scenario with two different values of the minimum Einstein radius: $0.5''$ and $1.0''$.
A minimum Einstein radius of $0.5''$ is close to what can be currently be achieved via visual inspection of high-resolution space-based images \cite{Gar++22}, while the $1.0''$ limit can be seen as a more conservative case.
\begin{figure*}
\includegraphics[width=\textwidth]{lens_cornerplot.eps}
\caption{
Comparison bewteen the properties of lens samples and the general galaxy population. Distribution in foreground galaxy parameters.
\label{fig:lenspars}
}
\end{figure*}

The most striking difference between the samples is in the stellar mass, which is much larger in the lenses than in the general population. Lenses tend to also have a larger halo mass, a smaller half-light radius at fixed stellar mass, and a larger stellar population synthesis mismatch parameter. The distribution in ellipticity and inner dark matter slope of the lenses instead look very similar to that of the parent population.
%This outcome was expected, given the results of the analysis of \Sref{sect:indlenses}.
Additionally, we can see that the amplitude of the strong lensing bias appears to be always larger in the lens sample with the more restrictive selection on Einstein radius.

Qualitatively, the result shown in \Fref{fig:lenspars} matches our expectations from \Sref{sect:indlenses}.
In order to quantitatively assess the amplitude of the strong lensing bias in various lensing properties, we proceeded as follows.

\Fref{fig:sourcepars} shows the distribution in source properties.

\subsection{Galaxy-quasar lenses}
